Chapter 16: Particle-laden gravity currents: the lock-release slumping regime at the laboratory scale

### 2024-03-01 paper/main.tex
temp commit modifying introduction

% \documentclass[twocolumn]{article}
\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx, caption}
\usepackage{authblk}
\usepackage{textcomp}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{cuted}

@@

	Particle-laden gravity currents (PLGCs) are driven by the mass difference between a heavy fluid-particle mixture and a lighter ambient liquid. They often occur in natural and industrial situations, among which a typical situation is the release of a finite volume. Here, we focus on such `dam-break' situations, which are studied using lock-release devices at the laboratory scale.
	%
	The objective of the present chapter is to provide a description at the macroscopic scale of the early moments of the flow, namely the \emph{slumping regime}, with respect to the relevant dimensionless parameters. For this, we combine a total of 288 runs from three different lock-release devices and from two-fluids numerical simulations, which allow us to cover a large range of particle types (size and density), volume fractions, bottom slopes and geometries.
	%
	By tracking the front propagation through time, we extract the dimensionless slumping velocity $\mathcal{F}r$ and dimensionless characteristic slumping duration $\tau$, on which we base our description.
	%
	Our results show that the slumping velocity increases with the bottom slope, but decreases with the particle volume fraction when the latter exceeds a critical value. However, it remains independent of particle settling processes, which on the other hand affects the slumping duration. Hence, above a critical threshold, $\tau$ decreases as the ratio between the settling velocity and characteristic current velocity increases. For all these regimes, we derive scalings and energetic balances that reproduce the observed trends. The latter comparison confirms the role of initial energy transfer from the initial state towards the slumping phase on the resulting dynamics. This initial process and its characterisation remain crucial to prescribe relevant initial conditions for large-scale predictive modelling.

@@

Density currents have been widely considered at the laboratory scale for their ubiquitous occurrence in geophysical flows \citep{Hopfinger1983,Simpson1999,Dufek2016}. As they are mostly associated with unsteady and non-uniform dynamics in nature, the lock-release device appeared to be a popular canonical configuration to mimic their behaviour at the laboratory scale \citep{Hacker1996,Shin2004,Nogueira2014}. In particular, studies often focused on characterising the front dynamics and the front shape as a macroscopic description of the underlying complex dynamics \citep{Rottman1983,Marino2005,Hogg2006}.

The basic approach to describe a lock-release evolution is built upon characteristics solutions of the hyperbolic shallow water equations for an inviscid dense liquid of density $\rho_{\text c}$ slumping on a horizontal bottom under an inertialess and weightless ambient fluid. In particular, this allows predicting the front velocity $U_{\text{c}}$ only based on the initial condition of the released mass. At the early stages of the slumping phase, it is found to be simply related to the only velocity scale of the system as
\begin{equation}
	U_{\text{c}}\propto \sqrt{gH_0},
\end{equation}
with $g$ gravitational acceleration and $H_0$ the initial height of the fluid volume being released. The method of characteristics suggests a coefficient of proportionality of exactly $2$ between velocities $U_{\text{c}}$ and $\sqrt{g H_0}$. An extension towards a weighted ambient fluid of density $\rho_{\text a}$ could lead to
\begin{equation}
	\label{eq:solUc}
	U_{\text{c}}=F\sqrt{\tilde{g}H_0},\quad\mbox{with}\quad \tilde{g}=\frac{\rho_{\text c}-\rho_{\text a}}{\rho_{\text c}}g,
\end{equation}
However, the coefficient of proportionality $F$, similar to a Froude number, now requires slightly more attention as it is clearly found smaller than $2$ in experiments or simulations. The main reason is that most density currents are characterized by a small density ratio between current and ambient, making the assumption of inertialess ambient fluid unlikely. A dynamic front condition is now required, and is often inspired from the steady front configuration derived by \citet{Benjamin1968}, now also linking the ambient height $H_{\text a}$ to the front velocity. This front relation includes two unknowns, a front velocity and a front height, and is often used as a front boundary condition in shallow water numerical solver for gravity currents \cite[e.g.][]{Ungarish2009}. Then, past studies managed to link the front velocity to the current height close to the front. However, the link between the front velocity and the initial conditions then becomes much more difficult to predict.
Models have been proposed to circumvent this issue, leading for instance to \citep{Huppert1980}
\begin{equation}
	\label{eq:Froude_Huppert1980}
	F = 0.5 \sqrt{\frac{\rho_{\text c}}{\rho_{\text a}}}\left(\frac{H_{\text a}}{H_0}\right)^{1/3}.
\end{equation}
%
The latter is relevant for not too deep ambient $H_{\text a}$ compared to the reservoir height $H_0$, which is actually the case in most experiments. Many configurations correspond to full-depth release ($H_{\text a}=H_0$), which leads to $F = 0.5(\rho_{\text c}/\rho_{\text a})^{1/2}$. Even if several experiments and numerical simulations confirm such a result, the exact value of $F$ remains still unclear and probably dependent on several physical processes. In particular, these relationships during the slumping regime are based on assuming negligible dissipation, which shall be violated in some cases, for instance in the presence of settling particles or significant mixing with the ambient at the miscible interface.

For a finite volume lock-release device on a mild enough slope, the slumping regime necessarily ends when the rarefaction wave associated with finite volume reaches the current front \citep[e.g.][]{Rottman1983, Hogg2006}. This leads to a new regime usually referred to as the self-similar regime in the literature. This is observed on the front evolution as $U_{\text{c}}$ is no longer time independent but evolves as $t^{-1/3}$.
A simple approach can be useful to provide a rapid estimation of the position $X_{\text t}$ or time $T_{\text t}$ at which this transition occurs. For that purpose, it is worth mentioning that such a transition occurs as a rarefaction wave evolves faster than the front. Then, assuming the front velocity $U_{\text{c}}$ to be different from the rarefaction wave velocity at the interface $\sqrt{\tilde{g}H_0}$, which is assumed to be constant along its entire propagation path for simplicity, this leads to a transition length scale $X_{\text t}$ written as
\begin{equation}
	\label{eq:Xt_inertial}
	\displaystyle X_{\text t}\propto  \frac{2L_0 F}{1 - F}.
\end{equation}
Note that here $2$ comes from the path of propagation of the wave: emanating from the gate, it propagates backwards to the end of the reservoir, reflects on the wall and then propagates towards the front.
During the slumping regime, the previous estimation of $F$ \eqref{eq:Froude_Huppert1980} for full depth release $H_0=H_{\text a}$ leads to $X_{\text t}\propto 2L_0/[2(\rho_{\text a}/\rho_{\text c})^{1/2}-1]$. Note that, although more advanced methods based on single-layer shallow-water models predict a similar result, $X_{\text t} = 2 L_0 F (1 + F/2)^{1/2}$ \citep{Hogg2006}, full depth-release experiments have found much larger transitional positions, i.e $X_{\text t} \approx 10 L_{0}$~\citep{Rottman1983}.

Other mechanisms can modify the dynamics of the slumping regime which could lead to other transitions of the front velocity at different positions $X_{\text t}$ or times $T_{\text t}$, prior to the influence of the rarefaction wave. This is basically part of the topic of the present chapter.

For a gravity current on a horizontal bottom, a classic transition discussed in the literature is the one dominated by viscous dissipation. The flow eventually reaches a fully viscous regime \citep{Huppert1980}. This transition typically holds for fully laminar flow. A similar regime can be induced by inertial friction, for which the full derivation remains less obvious \citep[see for instance][for the case of a constant-influx current]{Hogg2001}. In any case, such transition can be seen as the time or position at which the current is strongly affected by dissipation, which actually encompasses entrainment for miscible gravity currents. The slumping regime then ends when dissipation induced by bottom friction and/or entrainment at the miscible interface, overcomes the current dynamics.
%
Again using a simple idealised approach, the latter transition can be estimated from a dynamic condition balancing an estimation of the shear stress $\rho_{\text c} C_d U_{\text{c}}^2$ ($C_d$ being some drag coefficient induced by friction at the bottom and/or by entrainment at the interface) with inertia over the current height $h_{\text c}$, $\rho_{\text c} U_{\text{c}}^2 h_{\text c} /X_t$. In the latter expression, $h_{\text c}$ can have different scaling. In the case of a clearly detached current head, the constant height scales with the initial height $H_0$. When assuming that the current remains attached to the initial column at transition \citep[similarly to][but for inertial drag instead of viscous dissipation]{Bougouin2022}, mass conservation between the initial column and a simplified box geometry implies $h_{\text c} = H_0 L_0/X_t$.
%
Both cases respectively lead to
\begin{equation}
	\label{eq:Xt_drag}
	\displaystyle X_{\text t}= {\frac{H_0}{C_d}}\quad \mbox{or} \quad X_{\text t}=  \sqrt{\frac{H_0 L_0}{C_d}}.
\end{equation}

@@

Particle-Laden Gravity Currents (PLGCs) refer to as gravity currents, or density currents, are driven by a mass difference between a heavy fluid-particle mixture and a lighter ambient liquid \citep[e.g.][]{Hopfinger1983,Middleton1993,Kneller2000,Meiburg2010,wells2021}. Different subclasses of PLGCs can be distinguished depending on the relative density of the different phases constituting the system; i.e. the particle phase, the carrier fluid phase and the ambient fluid phase. In the following, we will focus on two of them, characterized by only two distinct densities. At one end, the carrier phase and ambient phase are the same fluid and particles are heavier. This is the most commonly studied subclass of PLGCs, usually referred to as turbidity currents. At the other end, PLGCs can also be induced by a neutrally-buoyant fluid-particle mixture flowing into a lighter ambient fluid.

A major difference between PLGCs and compositional currents is therefore the existence of typical length and time scales associated with the particle dynamics. Nevertheless, the prediction of these finite scales is difficult as it shall strongly depend on the interaction between particle dynamics and the local turbulence of the carrying fluid and/or ambient fluid at the upper interface which modifies entrainment and associated dissipation. In turn, this can modify the macroscopic behaviour of the current, for instance, the dynamics of the front. Accordingly, these dynamics shall be controlled by several dimensionless parameters that can lead to a wide variety of flow regimes. In particular, beyond the slope $\alpha$, the dimensionless density difference between the current and the ambient $\mathcal{A}t$, the ratio between available inertia and dissipation, say a Reynolds number $\mathcal{R}e$ as usually introduced as a control parameter for any density currents, the initial solid fraction of particles $\phi$ and a dimensionless settling velocity $S=V_{\text s}/U_{\text{c}}$ with $V_{\text s}$ the settling velocity of particles into the carrying fluid phase, have now to be considered as potentially affecting flow regimes. It shall be noted that in some specific situations, some of these dimensionless parameters could be linked (for instance $\phi$ and $\mathcal{A}t$ for heavy particles in a single-phase liquid for the carrying and ambient fluids).

Beyond the transition from the slumping regime to other regimes discussed in the previous section (still relevant for PLGCs although parametrisation of the processes involves more complex mechanisms induced by the fluid-particle interaction), here, a new transition towards the deposition of the particle can emerge. In particular, based on a kinematic approach balancing time scales related to current propagation and particle settling, leads to
%
\begin{equation}
	\label{eq:Xt_settling}
	X_{\text t} \propto H_0\frac{U_{\text{c}}}{V_{\text s}}.
\end{equation}
%
Note that such transition shall disappear for the neutrally-buoyant situation of PLGCs.
However, it has been observed that mixing at the interface can modify the local concentration of particles at the front, resulting in behaviour similar to that of settling~\cite{Schneider2023}. Again, such an approach only gives a scaling of transition, but the influence of the particle dynamics prior to such transition remains unclear. Overall, the value of $F$ in \eqref{eq:solUc} during the slumping regime remains to clarify varying the influence of the mechanisms discussed so far.

@@

A two-fluid model, SedFoam \citep{chauchat2017}, is used to simulate lock-release configurations from set-ups 1 and 2. This model available for download at Zenodo \citep{bonamy2023} is implemented in the open-source computational fluid dynamics toolbox OpenFOAM and has been extensively validated for various sediment transport problems such as sheet-flow \citep{cheng2016,chauchat2022,mathieu2021,mathieu2022}, scour \citep{nagel2020,tsai2022}, ripples \citep{salimi2021a,salimi2021b}, avalanches \citep{montella2021} and granular collapse \citep{montella2023}. The simulations presented herein are 2D vertical and turbulence-averaged. The turbulence model is a $k-\epsilon$ model with additional two-phase flow terms for density stratification and turbulent drag dissipation. The closure for the particle-particle interactions is a frictional-collisional model based on a Coulomb rheology for the frictional part and the kinetic theory of granular flows for the collisional part. The two contributions are simply added as described in \cite{chauchat2017}. The drag force is modelled using the \cite{schiller1933} closure multiplied by a hindrance function of \cite{richardson1954}. In order to describe the influence of the slope, the total pressure is solved and the gravity vector is rotated while the mesh is kept aligned with the $x$ and $y$ axis.

@@

269 experiments have been performed on the 3 experimental set-ups described in the previous section. They are complemented by 19 numerical simulations using SedFoam. Altogether, they are classified into 4 different datasets corresponding to the 3 experimental set-ups and the numerical modelling respectively. These 4 datasets are described below.

@@

The third dataset is gathered using setup 3. The experiments are performed using a neutrally-buoyant fluid-particle mixture made of polystyrene beads of typical diameter $d \sim 1~\textrm{mm}$ with density $\rho_{\text p} \sim 1038.5~\textrm{kg/m}^{3}$. Saltwater is added to the lock for density matching. The density difference between the lock and the ambient fluid is kept nearly constant for each experiment, equal to $25 \textrm{kg}~\textrm{m}^{-3}$. Only the initial volume fraction is varied in the range $[0.03$  $0.55]$.

@@

This dataset has been obtained using the two-fluid model described above. The domain size is $L_0+L_1=3.8$~m long and $W=0.5$~m height and the typical grid resolution is $\Delta x=1.25$~mm in the streamwise direction and it varies between $\Delta y \in [0.25 ; 3.5]$ mm in the wall-normal direction with a geometrical progression. The time step is adaptative following criteria on the Courant number, $C_{FL}=\max(U) \Delta t / \Delta x < 0.3$ for the phase velocities as well as for the relative velocity between the phases (the maximum allowed time-step during the simulation is set to $\max(\Delta t) = 10^{-3}$~s). The lock is $L_0=0.3$~m wide and $H_0=0.5$~m high, corresponding to the domain dimensions. The numerical schemes are second order in time and space: backward for time, a blend of centred and upwind schemes for advection (limitedLinear in OpenFOAM) and a centred scheme for diffusion. The initial condition is fluid and particle phase at rest ($u_f=u_p=0$) and the initial volume fraction of particles $\phi$ is set to a constant value in the region $x\in[-0.5 ; 0]$ m and $y\in[0;0.5]m$. At $t=0$, the governing equations are numerically integrated until $t_{end}=80$s maximum. The front position is determined as the most downstream position of the isocontour of volume fraction $\phi =10^{-3}$ at a frequency of 10Hz. This criterion has been demonstrated to be robust for all the simulations presented in this work. The particle properties are $e=0.8$ (restitution coefficient for binary collisions) and $\mu_{\text s}=0.53$ (macroscopic friction coefficient). Two sets of simulations have been performed. One mimics silica sand beads in water $\rho_{\text p}/\rho_{\text f}=2.65$ with diameter $d_p=60~\mu$m and initial volume fractions $\phi\in[0.015, 0.1]$. The other one mimics PMMA particles in water $\rho_{\text p}/\rho_{\text f}=1.19$ with diameters $d \in [100, 650]~\mu$m and initial volume fractions $\phi \in [0.0144, 0.5]$. For both configurations, the slope angle $\alpha$ has been varied in the range $\alpha \in [0^\circ, 45^\circ]$.

@@

%%%%%%%%%%%%%%%%%%%% Figure 3
\begin{figure*}[ht]
	\centering
	\includegraphics{figure3.pdf}
	\caption{(a) Current front position as a function of time for all experimental runs (b) Selected runs plotted in non-dimensional coordinates.}
	\label{fig:fig3}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

The current dynamics is characterized by the evolution of several dimensionless variables. Here, as stated in section~\ref{sec:intro}, we aim at giving a macroscopic description of the initial current dynamics across ranges of several parameters spanning multiple orders of magnitude. The wide diversity of experiments and simulations results in a large variety of currents which does not allow for an exhaustive description of their individual dynamics. We rather choose to focus on a single variable, the front position $x_{\text{f}}$, which has been shown in the literature to be a good proxy to describe the current dynamics~\citep[e.g.][]{bonnecaze1993particle, Chowdhury2011, Adduce2012}.

Figure~\ref{fig:fig3}(a) shows the front position of all experiments and simulations in dimensional coordinates, and figure~\ref{fig:fig3}(b) exhibits some selected runs in dimensionless coordinates. After a short transient acceleration phase, most currents exhibit at first a constant velocity phase, which corresponds to the so-called \emph{slumping} regime discussed in section \ref{sec:intro}. Afterwards, the velocity becomes time-dependent, often decreasing, due to one or multiple of the attenuation processes described in section~\ref{sec:intro}.
%
Qualitatively, it can be seen that increasing the bottom slope increases the slumping velocity while increasing the Stokes number shortens the duration of the constant-velocity phase (see fig~\ref{fig:fig3}).

>>>>

% \documentclass[twocolumn]{article}
\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx, caption}
\usepackage{authblk}
\usepackage{textcomp}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{cuted}

@@

	Particle-laden gravity currents (PLGCs) are driven by the mass difference between a heavy fluid-particle mixture and a lighter ambient liquid. They often occur in natural and industrial situations, among which a typical situation is the release of a finite volume. Here, we focus on such `dam-break' situations, which are studied using lock-release devices at the laboratory scale.
	%
	The objective of the present chapter is to describe the macroscopic scale of the early moments of the flow, namely the \emph{slumping regime}, with respect to the relevant dimensionless parameters. For this, we combine a total of 288 runs from three different lock-release devices and from two-fluids numerical simulations, which allow us to cover a large range of particle types (size and density), volume fractions, bottom slopes and geometries.
	%
	By tracking the front propagation through time, we extract the dimensionless slumping velocity $\mathcal{F}r$ and dimensionless characteristic slumping duration $\tau$, on which we base our description.
	%
	Our results show that the slumping velocity increases with the bottom slope, but decreases with the particle volume fraction when the latter exceeds a critical value. However, it remains independent of particle settling processes, which on the other hand affects the slumping duration. Hence, above a critical threshold, $\tau$ decreases as the ratio between the settling velocity and characteristic current velocity increases. For all these regimes, we derive scalings and energetic balances that reproduce the observed trends. The latter comparison confirms the role of initial energy transfer from the initial state towards the slumping phase on the resulting dynamics. This initial process and its characterization remain crucial to prescribe relevant initial conditions for large-scale predictive modeling.

@@

Density currents have been widely considered at the laboratory scale for their ubiquitous occurrence in geophysical flows \citep{Hopfinger1983,Simpson1999,Dufek2016}. As they are mostly associated with unsteady and non-uniform dynamics in nature, the lock-release device appeared to be a popular canonical configuration to mimic their behavior at the laboratory scale \citep{Hacker1996,Shin2004,Nogueira2014}. In particular, studies often focused on characterising the front dynamics and the front shape as a macroscopic description of the underlying complex dynamics \citep{Rottman1983,Marino2005,Hogg2006}.

The basic approach to describe a lock-release evolution is built upon characteristics solutions of the hyperbolic shallow water equations for an inviscid dense liquid of density $\rho_{\text c}$ slumping on a horizontal bottom under an inertialess and weightless ambient fluid. In particular, this allows predicting the front velocity $U_{\text{c}}$ only based on the initial condition of the released mass. At the early stages of the slumping phase, it is found to be simply related to the only velocity scale of the system as
\begin{equation}
	U_{\text{c}}\propto \sqrt{gH_0},
\end{equation}
with $g$ gravitational acceleration and $H_0$ the initial height of the fluid volume being released. The method of characteristics suggests a coefficient of proportionality of exactly $2$ between velocities $U_{\text{c}}$ and $\sqrt{g H_0}$. An extension towards a weighted ambient fluid of density $\rho_{\text a}$ could lead to
\begin{equation}
	\label{eq:solUc}
	U_{\text{c}}=F\sqrt{\tilde{g}H_0},\quad\mbox{with}\quad \tilde{g}=\frac{\rho_{\text c}-\rho_{\text a}}{\rho_{\text c}}g,
\end{equation}
However, the coefficient of proportionality $F$, similar to a Froude number, now requires slightly more attention as it is clearly found smaller than $2$ in experiments or simulations. The main reason is that most density currents are characterized by a small density ratio between current and ambient, making the assumption of inertialess ambient fluid unlikely. 
%
Hence, a dynamic front condition is now required, typically inspired by the steady front configuration derived by \citet{Benjamin1968}, also involving the ambient height $H_{\text a}$. This front relation includes two unknowns, a front velocity and a front height, and is used as a front boundary condition in the depth-averaged modeling of gravity currents \cite[e.g.][]{ungarish2020gravity}. Then, past studies managed to link the front velocity to the current height close to the front. However, the link between the front velocity and the initial conditions then becomes much more difficult to predict.
Models have been proposed to circumvent this issue, leading for instance to \citep{Huppert1980}
\begin{equation}
	\label{eq:Froude_Huppert1980}
	F = 0.5 \sqrt{\frac{\rho_{\text c}}{\rho_{\text a}}}\left(\frac{H_{\text a}}{H_0}\right)^{-1/3}.
\end{equation}
%
The latter is relevant only when typically $H_0/H_{\text a} > 0.2$, which is the case in most experiments. Many configurations correspond to full-depth release ($H_{\text a}=H_0$), which leads to $F = 0.5(\rho_{\text c}/\rho_{\text a})^{1/2}$. Even if several experiments and numerical simulations confirm such a result, the exact value of $F$ remains still unclear and probably dependent on several physical processes. In particular, these relationships during the slumping regime are based on assuming negligible dissipation, which shall be violated in some cases, for instance in the presence of settling particles or significant mixing with the ambient at the miscible interface.

For a finite volume lock-release device, the opening of the gate also generates a backward propagating rarefaction wave or bore (depending on $H_{0}/H_{\rm a}$), which reflects on the wall at the end of the reservoir and then propagates towards the front~\citep[e.g.][]{ungarish2020gravity}. As it evolves faster than the current front, it eventually reaches it, ending the slumping regime and leading to a new regime usually referred to as the self-similar regime~\citep[e.g.][]{Rottman1983, Hogg2006}. This is observed on the front evolution as $U_{\text{c}}$ is no longer time independent but evolves as $t^{-1/3}$.
%
Experiments, numerical simulations and theoretical studies using shallow-water modeling have shown that the position $X_{\text t}$ at which this transition occurs typically writes as:
%%
\begin{equation}
	\label{eq:Xt_inertial}
	\displaystyle X_{\text t} = \mathcal{A} L_{0},
\end{equation}
%%
where $\mathcal{A}$ is an increasing function of $H_{0}/H_{\rm a}$, and may depend on other parameters of the problem, e.g. $\rho_{\rm c}/\rho_{a}$. In the general case, its determination requires the numerical resolution of the two-layer shallow-water system of equations~\citep{Ungarish2005}. In the limit of deep currents (typically $H_{0}/H_{\rm a} \lesssim 0.25$), the system can be approximated by the one-layer shallow-water model, for which an analytical solution is available, $X_{\text t} = 2 L_0 F (1 + F/2)^{1/2}$ \citep{Hogg2006}. In the present case of full-depth release ($H_{0} = H_{\rm a}$), experiments and the resolution of the two-layer shallow-water model lead to $X_{\text t} \approx 8 \text{--} 12 L_{0}$~\citep{Rottman1983, Ungarish2005}.

Other mechanisms can modify the dynamics of the slumping regime, possibly leading to other possible positions $X_{\text t}$ or times $T_{\text t}$ at which the velocity is no longer time-independent. Depending on the relative intensity of these mechanisms, this can occur prior to the influence of the rarefaction wave or bore. Hence, determining the various regimes related to the prevalence of each mechanism is part of the topic of the present chapter.

For a gravity current on a horizontal bottom, a classic transition discussed in the literature is the one dominated by viscous dissipation. The flow eventually reaches a fully viscous regime \citep{Huppert1980}. This transition typically holds for fully laminar flow. A similar regime can be induced by inertial friction, for which the full derivation remains less obvious \citep[see for instance][for the case of a constant-influx current]{Hogg2001}. In any case, such transition can be seen as the time or position at which the current is strongly affected by dissipation, which encompasses entrainment for miscible gravity currents. The slumping regime then ends when dissipation induced by bottom friction and/or entrainment at the miscible interface, overcomes the current dynamics.
%
Again using a simple idealized approach, the latter transition can be estimated from a dynamic condition balancing an estimation of the shear stress $\rho_{\text c} C_d U_{\text{c}}^2$ ($C_d$ being some drag coefficient induced by friction at the bottom and/or by entrainment at the interface) with inertia over the current height $h_{\text c}$, $\rho_{\text c} U_{\text{c}}^2 h_{\text c} /X_t$. In the latter expression, $h_{\text c}$ can have different scalings. In the case of a detached current head, the constant height scales with the initial height $H_0$. When assuming that the current remains attached to the initial column at transition \citep[similar to][but for inertial drag instead of viscous dissipation]{Bougouin2022}, mass conservation between the initial column and a simplified box geometry implies $h_{\text c} = H_0 L_0/X_t$.
%
Both cases respectively lead to
\begin{equation}
	\label{eq:Xt_drag}
	\displaystyle X_{\text t}= {\frac{H_0}{C_d}}\quad \mbox{or} \quad X_{\text t}=  \sqrt{\frac{H_0 L_0}{C_d}}.
\end{equation}

@@

Particle-Laden Gravity Currents (PLGCs), also referred to as gravity currents or density currents, are driven by a mass difference between a heavy fluid-particle mixture and a lighter ambient liquid \citep[e.g.][]{Hopfinger1983, Middleton1993,Kneller2000, Meiburg2010,wells2021}. Different subclasses of PLGCs can be distinguished depending on the relative density of the different phases constituting the system; i.e. the particle phase, the carrier fluid phase and the ambient fluid phase. In the following, we will focus on two of them, characterized by only two distinct densities. At one end, the carrier phase and ambient phase are the same fluid and particles are heavier. This is the most commonly studied subclass of PLGCs, usually referred to as turbidity currents. At the other end, PLGCs can also be induced by a neutrally buoyant fluid-particle mixture flowing into a lighter ambient fluid.

A major difference between PLGCs and compositional currents is therefore the existence of typical length and time scales associated with the particle dynamics. Nevertheless, the prediction of these finite scales is difficult as it should strongly depend on the interaction between particle dynamics and the local turbulence of the carrying fluid and/or ambient fluid at the upper interface which modifies entrainment and associated dissipation. In turn, this can modify the macroscopic behavior of the current, for instance, the dynamics of the front. Accordingly, these dynamics shall be controlled by several dimensionless parameters that can lead to a wide variety of flow regimes. In particular, beyond the slope $\alpha$, the dimensionless density difference between the current and the ambient $\mathcal{A}t$, the ratio between available inertia and dissipation, say a Reynolds number $\mathcal{R}e$ as usually introduced as a control parameter for any density currents, the initial solid fraction of particles $\phi$ and a dimensionless settling velocity $S=V_{\text s}/U_{\text{c}}$ with $V_{\text s}$ the settling velocity of particles into the carrying fluid phase, have now to be considered as potentially affecting flow regimes. It shall be noted that in some specific situations, some of these dimensionless parameters could be linked (for instance $\phi$ and $\mathcal{A}t$ for heavy particles in a single-phase liquid for the carrying and ambient fluids).

Beyond the transition from the slumping regime to other regimes discussed in the previous section (still relevant for PLGCs although parametrization of the processes involves more complex mechanisms induced by the fluid-particle interaction), here, a new transition towards the deposition of the particle can emerge. In particular, based on a kinematic approach balancing time scales related to current propagation and particle settling, leads to
%
\begin{equation}
	\label{eq:Xt_settling}
	X_{\text t} \propto H_0\frac{U_{\text{c}}}{V_{\text s}}.
\end{equation}
%
Note that such transition shall disappear for the neutrally buoyant situation of PLGCs.
However, it has been observed that mixing at the interface can modify the local concentration of particles at the front, resulting in behavior similar to that of settling~\cite{Schneider2023}. Again, such an approach only gives a scaling of transition, but the influence of the particle dynamics prior to such transition remains unclear. Overall, the value of $F$ in \eqref{eq:solUc} during the slumping regime remains to clarify varying the influence of the mechanisms discussed so far.

@@

A two-fluid model, SedFoam \citep{chauchat2017}, is used to simulate lock-release configurations from set-ups 1 and 2. This model available for download at Zenodo \citep{bonamy2023} is implemented in the open-source computational fluid dynamics toolbox OpenFOAM and has been extensively validated for various sediment transport problems such as sheet-flow \citep{cheng2016,chauchat2022,mathieu2021,mathieu2022}, scour \citep{nagel2020,tsai2022}, ripples \citep{salimi2021a,salimi2021b}, avalanches \citep{montella2021} and granular collapse \citep{montella2023}. The simulations presented herein are 2D vertical and turbulence-averaged. The turbulence model is a $k-\epsilon$ model with additional two-phase flow terms for density stratification and turbulent drag dissipation. The closure for the particle-particle interactions is a frictional-collisional model based on a Coulomb rheology for the frictional part and the kinetic theory of granular flows for the collisional part. The two contributions are simply added as described in \cite{chauchat2017}. The drag force is modeled using the \cite{schiller1933} closure multiplied by a hindrance function of \cite{richardson1954}. In order to describe the influence of the slope, the total pressure is solved and the gravity vector is rotated while the mesh is kept aligned with the $x$ and $y$ axis.

@@

269 experiments have been performed on the 3 experimental set-ups described in the previous section. They are complemented by 19 numerical simulations using SedFoam. Altogether, they are classified into 4 different datasets corresponding to the 3 experimental set-ups and the numerical modeling respectively. These 4 datasets are described below.

@@

The third dataset is gathered using setup 3. The experiments are performed using a neutrally buoyant fluid-particle mixture made of polystyrene beads of typical diameter $d \sim 1~\textrm{mm}$ with density $\rho_{\text p} \sim 1038.5~\textrm{kg/m}^{3}$. Saltwater is added to the lock for density matching. The density difference between the lock and the ambient fluid is kept nearly constant for each experiment, equal to $25 \textrm{kg}~\textrm{m}^{-3}$. Only the initial volume fraction is varied in the range $[0.03$  $0.55]$.

@@

This dataset has been obtained using the two-fluid model described above. The domain size is $L_0+L_1=3.8$~m long and $W=0.5$~m height and the typical grid resolution is $\Delta x=1.25$~mm in the streamwise direction and it varies between $\Delta y \in [0.25; 3.5]$ mm in the wall-normal direction with a geometrical progression. The time step is adaptative following criteria on the Courant number, $C_{FL}=\max(U) \Delta t / \Delta x < 0.3$ for the phase velocities as well as for the relative velocity between the phases (the maximum allowed time-step during the simulation is set to $\max(\Delta t) = 10^{-3}$~s). The lock is $L_0=0.3$~m wide and $H_0=0.5$~m high, corresponding to the domain dimensions. The numerical schemes are second order in time and space: backward for time, a blend of centered and upwind schemes for advection (limitedLinear in OpenFOAM) and a centered scheme for diffusion. The initial condition is fluid and particle phase at rest ($u_f=u_p=0$) and the initial volume fraction of particles $\phi$ is set to a constant value in the region $x\in[-0.5 ; 0]$ m and $y\in[0;0.5]m$. At $t=0$, the governing equations are numerically integrated until $t_{end}=80$s maximum. The front position is determined as the most downstream position of the isocontour of volume fraction $\phi =10^{-3}$ at a frequency of 10Hz. This criterion has been demonstrated to be robust for all the simulations presented in this work. The particle properties are $e=0.8$ (restitution coefficient for binary collisions) and $\mu_{\text s}=0.53$ (macroscopic friction coefficient). Two sets of simulations have been performed. One mimics silica sand beads in water $\rho_{\text p}/\rho_{\text f}=2.65$ with diameter $d_p=60~\mu$m and initial volume fractions $\phi\in[0.015, 0.1]$. The other one mimics PMMA particles in water $\rho_{\text p}/\rho_{\text f}=1.19$ with diameters $d \in [100, 650]~\mu$m and initial volume fractions $\phi \in [0.0144, 0.5]$. For both configurations, the slope angle $\alpha$ has been varied in the range $\alpha \in [0^\circ, 45^\circ]$.

@@

%%%%%%%%%%%%%%%%%%%% Figure 3
\begin{figure*}[ht]
	\centering
	\includegraphics{figure3.pdf}
	\caption{(a) Dimensionless current front position as a function of dimensionless time for all experimental runs (b) Zoom on the early dynamics for selected runs. Here, $\alpha$ and $\mathcal{S}t$ increase along the black arrows.}
	\label{fig:fig3}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%

The current dynamics is characterized by the evolution of several dimensionless variables. Here, as stated in section~\ref{sec:intro}, we aim to give a macroscopic description of the initial current dynamics across ranges of several parameters spanning multiple orders of magnitude. The wide diversity of experiments and simulations results in a large variety of currents which does not allow for an exhaustive description of their individual dynamics. We rather choose to focus on a single variable, the front position $x_{\text{f}}$, which has been shown in the literature to be a good proxy to describe the current dynamics~\citep[e.g.][]{bonnecaze1993particle, Chowdhury2011, Adduce2012}.

Figure~\ref{fig:fig3}(a) shows the front position of all experiments and simulations in dimensional coordinates, and figure~\ref{fig:fig3}(b) exhibits some selected runs in dimensionless coordinates. After a short transient acceleration phase, most currents exhibit at first a constant velocity phase, which corresponds to the so-called \emph{slumping} regime discussed in section \ref{sec:intro}. Afterward, the velocity becomes time-dependent, often decreasing, due to one or multiple of the attenuation processes described in section~\ref{sec:intro}.
%
Qualitatively, it can be seen that increasing the bottom slope increases the slumping velocity while increasing the Stokes number shortens the duration of the constant-velocity phase (see fig~\ref{fig:fig3}).

In order to properly quantify the slumping regime, i.e. the associated front velocity and duration, it is necessary to prescribe the relevant attenuation process leading to the regime transition. Yet, this is a priori unknown given the variety of involved processes in the wide range of PLGCs parameters used here. We thus propose to overcome this issue by fitting experimental and numerical front evolution by a second-order temporal expansion of the front position $x_{\text f}(t)$ as
\begin{equation}
	\label{eq:fit_eq}
	\frac{x_{\text f}}{L_{0}} = \frac{x_{i}}{L_0} + \mathcal{F}r \left[\frac{t}{t_{0}} - \frac{1}{\tau} \left(\frac{t}{t_{0}}\right)^{2}\right],
\end{equation}
where $\mathcal{F}r = U_{\text{c}}/U_{0}$ corresponds to the dimensionless current velocity during the slumping regime and $\tau$ to a dimensionless attenuation time scale related to $X_{\text{t}}$ as
\begin{equation}
	\label{eq:link_Xt_tau}
	\frac{X_{\text{t}}}{L_{0}} \propto \tau\mathcal{F}r.
\end{equation}
Note that the previous expansion \eqref{eq:fit_eq} assumes a constant velocity and constant deceleration trend of the front evolution. The latter is relevant to extract the time scale of the existence of the slumping regime, i.e. the linear component, as long as its dynamics has only weakly deviated from such linear evolution. It is thus required to consider such a solution only during a brief time scale beyond the linear slumping regime.
%as defined here, $\mathcal{F}r$ is not a control parameter, but a dimensionless variable characteristic of the initial current dynamics (The Froude number which would be defined as a control parameter for a lock-release configuration would be $\mathcal{F}r \equiv U_{0}/U_{0} = 1$, and is thus not relevant to describe the flow dynamics).
